---
title: "EDA: Graph-based Features from Spatial Cell Data"
author: "Sehyun Oh"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteIndexEntry{Graph Features EDA}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
output:
  BiocStyle::html_document:
    number_sections: yes
    toc: yes
    toc_depth: 4
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(comment = "#>", 
                      collapse = TRUE, 
                      message = FALSE, 
                      warning = FALSE,
                      fig.width = 10,
                      fig.height = 7)
```

# Overview

This vignette demonstrates exploratory data analysis (EDA) of graph-based 
spatial features extracted from histopathology images. The data includes:

- Graph summary statistics (degree, mixing index)
- Cell type proportions and edge fractions
- Multiscale graph features at different radii (r20, r40, r80, r100)
- Smoothed embeddings from graph structure
- Cell-level features with graph properties

## Introduction to Graph-based Spatial Analysis

### What is Graph-based Analysis?

In spatial biology, tissue can be represented as a **graph (or network)** where:

- **Nodes** = Individual cells (identified through segmentation)
- **Edges** = Spatial relationships between cells (e.g., cells within a certain distance)
- **Graph properties** = Quantitative metrics describing tissue organization

This approach transforms complex spatial patterns into computable features that
can be analyzed statistically and used for machine learning.

### Why Use Graph-based Features?

1. **Captures Spatial Context**: Unlike traditional cell counting, graphs encode
   how cells are organized relative to each other
2. **Biologically Meaningful**: Cell-cell interactions drive tumor progression,
   immune responses, and treatment outcomes
3. **Quantifiable**: Provides objective, reproducible metrics
4. **Scalable**: Can analyze millions of cells across whole slide images
5. **Multi-scale**: Examines organization at different spatial scales

### Clinical Relevance

Spatial organization patterns are associated with:

- **Prognosis**: Survival outcomes in multiple cancer types
- **Treatment Response**: Immunotherapy and targeted therapy efficacy
- **Tumor Biology**: Invasion patterns, metastatic potential
- **Immune Function**: T cell exhaustion, immune evasion mechanisms

# Load Required Libraries

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(patchwork)
library(pheatmap)
library(corrplot)
library(RColorBrewer)
library(ggridges)
library(GGally)
library(curatedTCGAData)
library(kableExtra)
library(readr)
```

# Terminology and Definitions

## Key Graph Metrics

### Degree

**Definition**: The number of edges connected to a node (number of neighboring cells).

- **Mean Degree**: Average number of neighbors per cell
- **Higher values**: Denser cellular packing
- **Lower values**: Sparse cell distribution
- **Interpretation**: Reflects local cell density and tissue compactness

### Mixing Index

**Definition**: Measure of cell type diversity in local neighborhoods.

- **Range**: 0 to 1
- **Value = 0**: All neighbors are the same cell type (complete segregation)
- **Value = 1**: Perfectly uniform distribution of all cell types (complete mixing)
- **Value = 0.5**: Moderate mixing
- **Interpretation**: Quantifies spatial clustering vs. intermixing of cell types

**Biological Significance**:

- High mixing (>0.7): Diverse microenvironments, active cell-cell interactions
- Low mixing (<0.3): Segregated compartments, limited cross-talk
- Intermediate mixing: Partial compartmentalization with interface regions

### Edge Fraction

**Definition**: Proportion of edges connecting two specific cell types.

- **Formula**: (# edges between type A and B) / (total # edges)
- **Diagonal values**: Homotypic interactions (same cell type)
- **Off-diagonal values**: Heterotypic interactions (different cell types)
- **Interpretation**: Reveals preferential associations or spatial exclusion

## Cell Type Definitions

### Neoplastic Cells

- **Definition**: Malignant tumor cells
- **Characteristics**: Atypical morphology, irregular nuclei
- **Relevance**: Primary target for therapy, driver of disease

### Inflammatory Cells

- **Definition**: Immune cells (lymphocytes, macrophages, etc.)
- **Function**: Anti-tumor immunity or pro-tumor inflammation
- **Clinical Impact**: Predictive biomarker for immunotherapy

### Stromal Cells

- **Definition**: Fibroblasts, endothelial cells, support cells
- **Function**: ECM production, angiogenesis, structural support
- **Impact**: Can create physical barriers or promote invasion

### Benign Epithelial

- **Definition**: Normal epithelial cells
- **Location**: Adjacent normal tissue
- **Significance**: Reference for comparison to tumor

### Necrotic Regions

- **Definition**: Areas of cell death
- **Causes**: Hypoxia, nutrient deprivation
- **Implication**: Poor vascularization, aggressive disease

## Spatial Scales

Multiscale analysis examines features at different radii:

- **r20 (20 μm)**: Immediate microenvironment (~2-3 cell diameters)
  - Direct cell-cell contact
  - Juxtacrine signaling
  
- **r40 (40 μm)**: Local neighborhood (~5-6 cell diameters)
  - Paracrine signaling range
  - Microenvironmental niche
  
- **r80 (80 μm)**: Tissue compartment (~10-12 cell diameters)
  - Functional tissue units
  - Structural organization
  
- **r100 (100 μm)**: Broader region (~15 cell diameters)
  - Tissue-level architecture
  - Compartment boundaries

# Import Data
## Load clinical metadata
```{r}
pc <- curatedTCGAData("PRAD", "RNAseq2Gene", version = "2.1.1", dry.run = FALSE)
meta <- colData(pc)
```


## Load Graph Features Summary

```{r load-summary}
datDir <- system.file("extdata", package = "CellGraphR")

# Graph features summary for single sample
graph_summary <- read_csv(file.path(datDir, 
  "TCGA-HI-7168-01Z-00-DX1.e4aabefc-b633-4f54-ae92-a839f2d80158_graph_features_summary.csv"))

# Preview the data
glimpse(graph_summary)
```

The graph summary contains:
- Mean/std degree and mixing index
- Proportion of each cell type
- Edge fractions between all pairs of cell types

## Load Multiscale Graph Features

```{r load-multiscale}
# Multiscale features across multiple samples
multiscale_features <- read_csv(file.path(datDir, 
  "tcga_multiscale_graph_features_per_slide.csv"))

# Preview
glimpse(multiscale_features)
```

This dataset contains mean degree and mixing index for each cell type at 
different spatial scales (r20, r40, r80, r100 μm).

## Load Graph Smooth Embedding

```{r load-embedding}
# Smoothed embedding from graph structure
graph_embedding <- read_csv(file.path(datDir, 
  "TCGA-HI-7168-01Z-00-DX1.e4aabefc-b633-4f54-ae92-a839f2d80158_graph_smooth_embedding.csv"))

# Preview
glimpse(graph_embedding)
```

## Load Cell-level Graph Features (h5ad)

```{r load-h5ad, eval=FALSE}
# Load h5ad file containing cell-level graph features
# This requires the h5adToSe function from HistoImageR
library(HistoImageR)

h5ad_file <- file.path(datDir, 
  "TCGA-HI-7168-01Z-00-DX1.e4aabefc-b633-4f54-ae92-a839f2d80158_with_graph_features.h5ad")

if (file.exists(h5ad_file)) {
  cell_graph_data <- h5adToSe(h5ad_file)
  print(cell_graph_data)
}
```

# EDA: Graph Summary Statistics

## Overall Graph Properties

```{r graph-properties}
# Extract key statistics
cat("Mean Degree:", graph_summary$mean_degree, "\n")
cat("Std Degree:", graph_summary$std_degree, "\n")
cat("Mean Mixing Index:", graph_summary$mean_mixing_index, "\n")
cat("Std Mixing Index:", graph_summary$std_mixing_index, "\n")
```

### Interpretation of Graph Statistics

**Mean Degree** (`r round(graph_summary$mean_degree, 2)`):

- Each cell has an average of ~`r round(graph_summary$mean_degree, 1)` neighbors
- This indicates **moderate cell density**
- Typical range: 2-6 neighbors in 2D tissue sections
- **Clinical Context**: Reflects tissue compactness and cellular organization

**Standard Deviation of Degree** (`r round(graph_summary$std_degree, 2)`):

- High variability (CV = `r round(graph_summary$std_degree/graph_summary$mean_degree*100, 1)`%) indicates **spatial heterogeneity**
- Some regions are densely packed, others are sparse
- **Biological Significance**: Heterogeneous density often associated with:
  - Tumor invasive fronts (dense tumor, sparse stroma)
  - Immune infiltration patterns (clustered vs. diffuse)
  - Necrotic regions (sparse cells)

**Mean Mixing Index** (`r round(graph_summary$mean_mixing_index, 2)`):

- Value of ~`r round(graph_summary$mean_mixing_index, 2)` indicates **partial mixing**
- Cell types are neither completely segregated nor uniformly mixed
- Suggests presence of both:
  - Homogeneous compartments (e.g., tumor nests, immune clusters)
  - Interface regions with diverse cell types
- **Therapeutic Implications**: Moderate mixing may facilitate immune-tumor interactions

**Standard Deviation of Mixing Index** (`r round(graph_summary$std_mixing_index, 2)`):

- High variability indicates **diverse microenvironmental niches**
- Some regions show strong clustering, others show mixing
- **Research Implications**: Spatial heterogeneity may contribute to:
  - Treatment resistance (sanctuary sites)
  - Clonal evolution
  - Varied immune responses

## Cell Type Composition

```{r cell-type-composition}
# Extract cell type proportions
cell_types <- c("benign epithelial", "inflammatory", "necrotic", 
                "neoplastic", "no label", "stromal")

type_props <- graph_summary %>%
  select(starts_with("prop_type_")) %>%
  pivot_longer(everything(), 
               names_to = "cell_type", 
               values_to = "proportion") %>%
  mutate(cell_type = gsub("prop_type_", "", cell_type))

# Visualization
ggplot(type_props, aes(x = reorder(cell_type, -proportion), 
                       y = proportion, fill = cell_type)) +
  geom_bar(stat = "identity") +
  scale_fill_brewer(palette = "Set3") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") +
  labs(title = "Cell Type Composition",
       x = "Cell Type",
       y = "Proportion") +
  scale_y_continuous(labels = scales::percent)
```

### Interpretation: Tumor Microenvironment Composition

```{r composition-interpretation}
# Calculate key metrics
inflam_pct <- round(type_props$proportion[type_props$cell_type == "inflammatory"] * 100, 1)
stromal_pct <- round(type_props$proportion[type_props$cell_type == "stromal"] * 100, 1)
neoplastic_pct <- round(type_props$proportion[type_props$cell_type == "neoplastic"] * 100, 1)
immune_stromal_ratio <- round(type_props$proportion[type_props$cell_type == "inflammatory"] / 
                               type_props$proportion[type_props$cell_type == "stromal"], 2)
```

**Key Observations**:

1. **Immune-Dominant Landscape** (`r inflam_pct`% inflammatory cells)
   - **High immune infiltration** suggests active immune response
   - May indicate "hot" tumor phenotype
   - **Clinical Implication**: Potentially responsive to immunotherapy
   - **Prognostic Value**: Often associated with better outcomes in many cancers

2. **Abundant Stroma** (`r stromal_pct`% stromal cells)
   - Extensive extracellular matrix and support structure
   - **Desmoplastic reaction**: Fibrotic response to tumor
   - **Dual Role**:
     - Physical barrier to immune cells and drugs
     - Source of growth factors and signaling molecules
   - **Therapeutic Target**: Stromal modulation may enhance drug delivery

3. **Low Neoplastic Fraction** (`r neoplastic_pct`% tumor cells)
   - **Possible Interpretations**:
     - Early stage disease with extensive reaction
     - Highly infiltrated tumor with immune rejection
     - Post-treatment sample with residual disease
     - Sampling from tumor periphery rather than core
   - **Important Note**: Low tumor content may affect molecular analyses (DNA/RNA)

4. **Immune-to-Stromal Ratio** (`r immune_stromal_ratio`:1)
   - Balanced immune-stromal compartments
   - Suggests complex TME with multiple cell-cell interactions
   - Ratio can predict treatment response in some cancer types

**Clinical Context**:

This TME composition profile (immune-high, stroma-high, tumor-low) is observed in:

- Immunogenic tumors with strong host response
- Response to prior therapy
- Certain histological subtypes (e.g., medullary carcinoma)
- Favorable prognostic subset in multiple cancer types

## Edge Fraction Heatmap

Visualize the connectivity between different cell types:

```{r edge-fraction-heatmap, fig.width=10, fig.height=8}
# Extract edge fractions
edge_fracs <- graph_summary %>%
  select(starts_with("edge_frac_")) %>%
  pivot_longer(everything(), 
               names_to = "edge_type", 
               values_to = "fraction") %>%
  mutate(
    # Remove "edge_frac_" prefix and split on remaining underscores
    edge_type = gsub("^edge_frac_", "", edge_type)
  ) %>%
  separate(edge_type, 
           into = c("from", "to"), 
           sep = "_(?=[^_]*$)",  # Split on last underscore
           extra = "merge") %>%
  mutate(
    from = gsub("_", " ", from),
    to = gsub("_", " ", to)
  )

# Create matrix for heatmap
edge_matrix <- edge_fracs %>%
  pivot_wider(names_from = to, values_from = fraction) %>%
  column_to_rownames("from") %>%
  as.matrix()

# Plot heatmap
pheatmap(edge_matrix,
         main = "Edge Fractions Between Cell Types",
         color = colorRampPalette(c("white", "navy"))(100),
         display_numbers = TRUE,
         number_format = "%.3f",
         cluster_rows = FALSE,
         cluster_cols = FALSE,
         fontsize = 10,
         angle_col = 45)
```

### Interpretation: Cell-Cell Interaction Patterns

**Reading the Heatmap**:

- **Rows**: Source cell type
- **Columns**: Target cell type (neighbor)
- **Values**: Fraction of all edges connecting these two types
- **Diagonal**: Homotypic interactions (same type neighbors)
- **Off-diagonal**: Heterotypic interactions (different type neighbors)

```{r edge-interpretation, echo=FALSE}
# Extract key interaction patterns
# Check if matrix has correct row/column names
if ("inflammatory" %in% rownames(edge_matrix)) {
  inflam_self <- edge_matrix["inflammatory", "inflammatory"]
  stromal_self <- edge_matrix["stromal", "stromal"]
  inflam_stromal <- edge_matrix["inflammatory", "stromal"]
  neoplastic_self <- edge_matrix["neoplastic", "neoplastic"]
  inflam_neoplastic <- edge_matrix["inflammatory", "neoplastic"]
} else {
  # Matrix might not be square or names might be different
  # Find the actual names
  cell_type_names <- rownames(edge_matrix)
  cat("Available cell types:", paste(cell_type_names, collapse=", "), "\n")
  
  # Use flexible matching
  inflam_idx <- grep("inflam", cell_type_names, ignore.case=TRUE)[1]
  stromal_idx <- grep("stromal", cell_type_names, ignore.case=TRUE)[1]
  neopl_idx <- grep("neopl", cell_type_names, ignore.case=TRUE)[1]
  
  inflam_self <- edge_matrix[inflam_idx, inflam_idx]
  stromal_self <- edge_matrix[stromal_idx, stromal_idx]
  inflam_stromal <- edge_matrix[inflam_idx, stromal_idx]
  neoplastic_self <- edge_matrix[neopl_idx, neopl_idx]
  inflam_neoplastic <- edge_matrix[inflam_idx, neopl_idx]
}
```

**Major Interaction Patterns**:

1. **Inflammatory Self-Clustering** (`r round(inflam_self*100, 1)`%)
   - **Highest edge fraction**: Inflammatory cells preferentially neighbor each other
   - **Biological Meaning**:
     - Formation of lymphoid aggregates or tertiary lymphoid structures (TLS)
     - Coordinated immune response
     - Potential immune sanctuaries
   - **Clinical Significance**: TLS associated with favorable prognosis and immunotherapy response

2. **Stromal Networks** (`r round(stromal_self*100, 1)`%)
   - Strong stromal-stromal connectivity
   - **Interpretation**: Fibrous tissue forming continuous networks
   - **Functional Impact**:
     - Creates architectural framework
     - May impede immune cell migration
     - Affects drug penetration

3. **Immune-Stromal Interface** (`r round(inflam_stromal*100, 1)`%)
   - Substantial inflammatory-stromal edges
   - **Biological Processes**:
     - Immune cells trafficking through stroma
     - Cytokine-mediated remodeling
     - Stromal-immune crosstalk
   - **Therapeutic Implications**: Target for combination therapies (immune + stromal modulation)

4. **Limited Tumor Cell Connectivity** (`r round(neoplastic_self*100, 1)`%)
   - Low neoplastic-neoplastic interactions
   - **Possible Explanations**:
     - Infiltrative growth pattern (single cells or small clusters)
     - Tumor cells dispersed among stromal/immune compartments
     - Early invasion with disseminated cells
   - **Clinical Relevance**: Infiltrative patterns often more aggressive

5. **Immune-Tumor Contact** (`r round(inflam_neoplastic*100, 1)`%)
   - Moderate inflammatory-neoplastic edges
   - **Interpretation**: Active immune surveillance
   - **Context-Dependent**:
     - Could be anti-tumor immunity (good)
     - Could be immunosuppressive environment (bad)
     - Requires functional immune profiling to distinguish

**Spatial Organization Summary**:

The edge pattern reveals a **compartmentalized yet interactive TME**:

- **Immune compartment**: Self-organizing clusters
- **Stromal compartment**: Continuous networks
- **Tumor cells**: Scattered, highly exposed to immune/stromal interactions
- **Interface zones**: Active immune-stromal and immune-tumor interactions

This architecture suggests potential for:

- Immune-mediated tumor control
- Therapeutic vulnerabilities at interface regions
- Stromal-mediated treatment resistance

# EDA: Multiscale Features

## Overview of Samples

```{r sample-overview}
cat("Number of samples:", nrow(multiscale_features), "\n")
cat("Number of features:", ncol(multiscale_features) - 1, "\n")
cat("\nSample IDs:\n")
print(multiscale_features$slide_id)
```

## Feature Distribution by Scale

Examine how features vary across different spatial scales:

```{r feature-by-scale, fig.width=12, fig.height=8}
# Extract degree features across scales
degree_data <- multiscale_features %>%
  select(slide_id, matches("mean_deg.*_r\\d+")) %>%
  pivot_longer(-slide_id, 
               names_to = "feature", 
               values_to = "degree") %>%
  separate(feature, 
           into = c("stat", "type1", "cell_type", "radius"),
           sep = "_",
           extra = "merge") %>%
  mutate(
    cell_type = gsub("_r\\d+$", "", cell_type),
    radius = as.numeric(gsub("r", "", radius))
  )

# Plot
ggplot(degree_data, 
       aes(x = factor(radius), y = degree, color = cell_type, group = cell_type)) +
  geom_line(stat = "summary", fun = "mean", size = 1) +
  geom_point(stat = "summary", fun = "mean", size = 3) +
  facet_wrap(~ slide_id, scales = "free_y", ncol = 2) +
  scale_color_brewer(palette = "Set2") +
  theme_bw() +
  theme(legend.position = "bottom") +
  labs(title = "Mean Degree by Cell Type Across Spatial Scales",
       x = "Radius (μm)",
       y = "Mean Degree",
       color = "Cell Type")
```

### Interpretation: Scale-Dependent Connectivity

**What Does Degree Tell Us?**

- **Degree = connectivity**: Number of cells within specified radius
- **Increases with radius**: Larger neighborhoods contain more cells
- **Cell type differences**: Reflect both abundance and spatial distribution

**Key Patterns to Observe**:

1. **Linear vs. Non-linear Scaling**:
   - **Linear increase**: Uniform cell distribution
   - **Non-linear (plateau)**: Boundaries or compartmentalization
   - **Rapid increase then plateau**: Dense local clusters, sparse between

2. **Cell Type Rankings**:
   - **Highest degree**: Most abundant cell types (inflammatory, stromal)
   - **Lowest degree**: Rare cell types (neoplastic, necrotic)
   - **Consistent ranking**: Maintained spatial hierarchy

3. **Inter-Sample Variability**:
   - **Similar patterns**: Conserved tissue architecture
   - **Different patterns**: Distinct TME phenotypes
   - **Scale-specific differences**: Organization at specific spatial scales

**Biological Insights**:

- **Inflammatory cells** (high degree at all scales):
  - Widely distributed throughout tissue
  - Form dense local clusters
  - Infiltrate multiple compartments

- **Stromal cells** (moderate, consistent degree):
  - Evenly distributed networks
  - Continuous tissue framework
  - Scale-independent organization

- **Neoplastic cells** (low degree):
  - Sparse distribution or small clusters
  - Limited expansion at larger scales
  - Suggests infiltrative rather than expansive growth

**Clinical Implications**:

- **Scale-dependent features** may capture different biological processes
- **20 μm**: Direct cell-cell contact (signaling)
- **40-80 μm**: Microenvironmental niches (metabolic gradients)
- **100 μm**: Tissue architecture (compartment organization)

## Mixing Index Across Scales

```{r mixing-index-scale, fig.width=12, fig.height=8}
# Extract mixing index features
mixing_data <- multiscale_features %>%
  select(slide_id, matches("mean_mix.*_r\\d+")) %>%
  pivot_longer(-slide_id, 
               names_to = "feature", 
               values_to = "mixing_index") %>%
  separate(feature, 
           into = c("stat1", "stat2", "cell_type", "radius"),
           sep = "_",
           extra = "merge") %>%
  mutate(
    cell_type = gsub("_r\\d+$", "", cell_type),
    radius = as.numeric(gsub("r", "", radius))
  )

# Plot
ggplot(mixing_data, 
       aes(x = factor(radius), y = mixing_index, color = cell_type, group = cell_type)) +
  geom_line(stat = "summary", fun = "mean", size = 1) +
  geom_point(stat = "summary", fun = "mean", size = 3) +
  facet_wrap(~ slide_id, scales = "free_y", ncol = 2) +
  scale_color_brewer(palette = "Set2") +
  theme_bw() +
  theme(legend.position = "bottom") +
  labs(title = "Mixing Index by Cell Type Across Spatial Scales",
       x = "Radius (μm)",
       y = "Mean Mixing Index",
       color = "Cell Type")
```

### Interpretation: Scale-Dependent Cell Type Mixing

**Understanding Mixing Index Trends**:

- **Increasing with scale**: Cell type becomes more intermixed at larger distances
  - Local clustering (same type) but regional diversity
  - Example: Tumor nests (low mixing locally) within diverse stroma (high mixing regionally)

- **Decreasing with scale**: Cell type segregation increases at larger distances
  - Compartmentalized tissue architecture
  - Distinct regions dominated by single cell type

- **Stable across scales**: Consistent mixing pattern
  - Uniform distribution at all scales
  - Well-mixed compartment

**Typical Patterns by Cell Type**:

1. **Inflammatory Cells** (often high, increasing mixing):
   - Can form local clusters (lymphoid aggregates)
   - But infiltrate broadly (high regional mixing)
   - Scale increase reflects transition from cluster to infiltration

2. **Stromal Cells** (moderate, stable mixing):
   - Distributed throughout tissue
   - Continuous network architecture
   - Consistent mixing reflects uniform distribution

3. **Neoplastic Cells** (low, variable mixing):
   - Often form tumor nests (low mixing locally)
   - May remain segregated at larger scales (invasive front)
   - Or become more mixed (infiltrative pattern)
   - Pattern reveals growth mechanism

**Clinical Significance**:

- **High mixing at small scales**: Direct cell-cell interactions
  - Immune-tumor contact (immunotherapy targets)
  - Stromal-tumor contact (invasion facilitators)

- **Low mixing at small scales**: Compartmentalization
  - Tumor sanctuary from immune surveillance
  - Limited therapeutic access

- **Scale-dependent transitions**: Hierarchical organization
  - Local homogeneity + regional heterogeneity
  - Multiple therapeutic vulnerabilities

**Research Applications**:

- **Biomarker Development**: Different scales predict different outcomes
- **Treatment Planning**: Target interfaces (high mixing) vs. compartments (low mixing)
- **Disease Monitoring**: Changes in mixing patterns reflect treatment response

## Heatmap of All Features

```{r feature-heatmap, fig.width=12, fig.height=10}
# Prepare data for heatmap
heatmap_data <- multiscale_features %>%
  column_to_rownames("slide_id") %>%
  as.matrix()

# Normalize by column (z-score)
heatmap_data_scaled <- scale(heatmap_data)

# Plot
pheatmap(heatmap_data_scaled,
         main = "Multiscale Graph Features (Z-scored)",
         color = colorRampPalette(c("blue", "white", "red"))(100),
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         clustering_method = "ward.D2",
         show_colnames = FALSE,
         fontsize = 10,
         angle_col = 45)
```

## Correlation Analysis

```{r correlation-analysis, fig.width=10, fig.height=10}
# Calculate correlation matrix
cor_matrix <- cor(multiscale_features %>% select(-slide_id), 
                  use = "pairwise.complete.obs")

# Plot correlation matrix (subset for visibility)
# Select every 5th feature to reduce clutter
idx <- seq(1, ncol(cor_matrix), by = 5)
corrplot(cor_matrix[idx, idx],
         method = "color",
         type = "upper",
         tl.col = "black",
         tl.cex = 0.6,
         title = "Feature Correlations (Subset)",
         mar = c(0, 0, 2, 0))
```

### Interpretation: Feature Correlations

**Understanding the Correlation Matrix**:

- **Blue**: Positive correlation (features increase together)
- **Red**: Negative correlation (one increases, other decreases)
- **White**: No correlation (features independent)
- **Intensity**: Strength of correlation

**What to Look For**:

1. **Blocks of High Correlation**:
   - Features that co-vary across samples
   - May represent:
     - Same biological process at different scales
     - Coordinated cellular programs
     - Redundant measurements (can reduce dimensionality)

2. **Negative Correlations**:
   - Trade-offs or mutual exclusion
   - Example: High tumor mixing negatively correlates with tumor segregation
   - Biological constraints (limited space, competitive processes)

3. **Uncorrelated Features**:
   - Independent aspects of tissue organization
   - Provide complementary information
   - Both needed for complete characterization

**Biological Examples**:

1. **Degree Features Across Scales** (positive correlation):
   - Samples with high degree at r20 also high at r100
   - Consistent density across scales
   - Reflects overall tissue cellularity

2. **Tumor vs. Immune Features** (may be negative):
   - High tumor presence, low immune infiltration (cold tumors)
   - Or vice versa (hot tumors)
   - Fundamental TME phenotype distinction

3. **Same Cell Type Across Metrics** (positive):
   - Cell type's degree correlates with its mixing
   - Abundant cell types have more neighbors AND mix more

**Applications**:

- **Feature Selection**: Remove highly correlated features (redundant)
- **Feature Engineering**: Combine correlated features into composite scores
- **Biological Insight**: Correlation structure reveals tissue organization principles
- **Quality Control**: Expected correlations validate data quality

## PCA Analysis

```{r pca-analysis, fig.width=12, fig.height=5}
# Perform PCA
pca_result <- prcomp(multiscale_features %>% select(-slide_id), 
                     center = TRUE, 
                     scale. = TRUE)

# Variance explained
var_explained <- summary(pca_result)$importance[2, ]

# PCA scores
pca_scores <- as.data.frame(pca_result$x) %>%
  mutate(slide_id = multiscale_features$slide_id)

# Plot
p1 <- ggplot(pca_scores, aes(x = PC1, y = PC2, label = slide_id)) +
  geom_point(size = 3, color = "steelblue") +
  geom_text(vjust = -0.5, hjust = 0.5, size = 2.5) +
  theme_bw() +
  labs(title = "PCA of Multiscale Graph Features",
       x = paste0("PC1 (", round(var_explained[1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(var_explained[2] * 100, 1), "%)"))

# Scree plot
p2 <- data.frame(
  PC = 1:min(10, length(var_explained)),
  Variance = var_explained[1:min(10, length(var_explained))]
) %>%
  ggplot(aes(x = PC, y = Variance)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_line(color = "red") +
  geom_point(color = "red", size = 2) +
  theme_bw() +
  labs(title = "Scree Plot",
       x = "Principal Component",
       y = "Proportion of Variance Explained") +
  scale_x_continuous(breaks = 1:10)

p1 + p2
```

### Interpretation: Principal Components Analysis

**What Does PCA Tell Us?**

PCA identifies the **major axes of variation** across samples:

- **PC1**: Primary source of difference between samples
- **PC2**: Secondary source of variation (orthogonal to PC1)
- **Variance explained**: How much information each PC captures

**Reading the Plots**:

**Scree Plot** (right):

- Shows variance explained by each PC
- **Steep drop**: First few PCs capture most variation
- **Elbow point**: Meaningful PCs vs. noise
- **Interpretation**: If first 2-3 PCs explain >60%, samples differ in few major ways

**PCA Score Plot** (left):

- Each point = one sample
- **Distance between points**: Similarity (close = similar, far = different)
- **Clusters**: Samples with similar spatial phenotypes
- **Outliers**: Unusual spatial organization

**Biological Meaning**:

1. **Separation along PC1**: 
   - Fundamental difference in tissue organization
   - Could reflect:
     - Disease stage (early vs. advanced)
     - Histological subtype
     - Treatment status (naive vs. treated)
     - Anatomical location

2. **Separation along PC2**:
   - Secondary source of variation
   - Often captures:
     - Immune infiltration patterns (hot vs. cold)
     - Stromal content (desmoplastic vs. cellular)
     - Tumor growth pattern (expansive vs. infiltrative)

3. **Clustering of samples**:
   - Identifies spatial phenotypes
   - Samples in same cluster have similar TME organization
   - Could define patient subgroups for precision medicine

**Clinical Applications**:

- **Patient Stratification**: PCA clusters may have different outcomes
- **Biomarker Discovery**: Features driving PC1/PC2 are most discriminating
- **Quality Control**: Outliers may indicate technical artifacts
- **Treatment Selection**: Different clusters may respond to different therapies

## Top Contributing Features to PC1 and PC2

```{r pca-loadings, fig.width=10, fig.height=6}
# Extract loadings
loadings <- as.data.frame(pca_result$rotation[, 1:2])
loadings$feature <- rownames(loadings)

# Top features for PC1
top_pc1 <- loadings %>%
  arrange(desc(abs(PC1))) %>%
  head(20)

# Top features for PC2
top_pc2 <- loadings %>%
  arrange(desc(abs(PC2))) %>%
  head(20)

# Plot
p1 <- ggplot(top_pc1, aes(x = reorder(feature, PC1), y = PC1)) +
  geom_bar(stat = "identity", fill = "coral") +
  coord_flip() +
  theme_bw() +
  theme(axis.text.y = element_text(size = 8)) +
  labs(title = "Top 20 Features Contributing to PC1",
       x = "", y = "Loading")

p2 <- ggplot(top_pc2, aes(x = reorder(feature, PC2), y = PC2)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +
  theme_bw() +
  theme(axis.text.y = element_text(size = 8)) +
  labs(title = "Top 20 Features Contributing to PC2",
       x = "", y = "Loading")

p1 / p2
```

### Interpretation: Feature Importance

**Understanding Loadings**:

- **Loadings** = contribution of each original feature to principal components
- **Magnitude** (absolute value): How much the feature contributes
- **Sign** (positive/negative): Direction of contribution
- **Top features**: Most important for distinguishing samples

**How to Interpret**:

1. **High Positive Loading on PC1**:
   - Samples with high PC1 scores have high values of this feature
   - Feature increases along PC1 axis

2. **High Negative Loading on PC1**:
   - Samples with high PC1 scores have low values of this feature
   - Feature decreases along PC1 axis

3. **Features with Similar Loadings**:
   - Tend to vary together across samples
   - May represent coordinated biological processes

**Biological Interpretation Strategy**:

1. **Identify top features**:
   - Which cell types?
   - Which spatial scales?
   - Degree or mixing metrics?

2. **Look for patterns**:
   - Do top features cluster by cell type?
   - Are specific scales over-represented?
   - Do homotypic or heterotypic interactions dominate?

3. **Generate hypotheses**:
   - PC1 dominated by inflammatory features → immune variation is primary axis
   - PC1 dominated by large-scale features → tissue architecture varies most
   - PC2 dominated by tumor features → tumor cell organization is secondary axis

**Example Interpretations**:

- **If inflammatory mixing at r100 is top feature**:
  - Major sample differences in large-scale immune distribution
  - Some samples have diffuse immune infiltration
  - Others have focal immune clusters
  - Clinical correlation: May predict immunotherapy response

- **If stromal degree at r20 is top feature**:
  - Samples differ in local stromal density
  - Desmoplastic vs. cellular stroma
  - Clinical correlation: May affect drug penetration

**Research Applications**:

- **Feature Selection**: Use top loading features for machine learning
- **Hypothesis Generation**: Features suggest biological processes to investigate
- **Validation**: Confirm importance in independent cohorts

# EDA: Graph Smooth Embedding

```{r embedding-analysis}
# Display embedding statistics
summary(graph_embedding)

# Check data structure
glimpse(graph_embedding)
```

The smoothed embedding represents aggregated features from the graph structure, 
capturing information about cell type composition, spatial arrangement, and 
intensity characteristics.

```{r embedding-visualization, fig.width=12, fig.height=8}
# Prepare data for visualization
embedding_long <- graph_embedding %>%
  select(-1) %>%  # Remove index column if present
  pivot_longer(everything(), 
               names_to = "feature", 
               values_to = "value")

# Create feature groups
embedding_long <- embedding_long %>%
  mutate(
    feature_group = case_when(
      str_detect(feature, "type_") ~ "Cell Type",
      str_detect(feature, "intensity") ~ "Intensity",
      str_detect(feature, "nearest_neighbor") ~ "Spatial",
      str_detect(feature, "degree") ~ "Graph Topology",
      str_detect(feature, "mixing") ~ "Graph Topology",
      TRUE ~ "Other"
    )
  )

# Visualize feature distributions
ggplot(embedding_long, aes(x = feature, y = value, fill = feature_group)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ feature_group, scales = "free", ncol = 2) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
        legend.position = "none") +
  labs(title = "Graph Smooth Embedding Features",
       x = "", y = "Value")
```

# Comparative Analysis

## Degree vs Mixing Index Relationship

```{r degree-mixing-relationship, fig.width=10, fig.height=6}
# Combine data for relationship analysis
degree_summary <- degree_data %>%
  group_by(slide_id, cell_type) %>%
  summarize(mean_degree = mean(degree, na.rm = TRUE), .groups = "drop")

mixing_summary <- mixing_data %>%
  group_by(slide_id, cell_type) %>%
  summarize(mean_mixing = mean(mixing_index, na.rm = TRUE), .groups = "drop")

combined <- left_join(degree_summary, mixing_summary, 
                     by = c("slide_id", "cell_type"))

# Plot
ggplot(combined, aes(x = mean_degree, y = mean_mixing, color = cell_type)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.2) +
  facet_wrap(~ cell_type, scales = "free") +
  scale_color_brewer(palette = "Set2") +
  theme_bw() +
  theme(legend.position = "none") +
  labs(title = "Degree vs Mixing Index Relationship by Cell Type",
       x = "Mean Degree (across scales)",
       y = "Mean Mixing Index (across scales)")
```

# Multi-modal Analysis

## Clinical Metadata Integration

### Load and Display Clinical Data

```{r load-clinical}
ids <- multiscale_features$slide_id %>% sub("^(TCGA-[0-9A-Z]{2}-[0-9A-Z]{4}).*", "\\1", .)
meta_sub <- meta[meta$patientID %in% ids,]

# Check available columns and create summary with safe column access
get_col_safe <- function(df, colname) {
  if (colname %in% colnames(df)) {
    return(df[[colname]])
  } else {
    return(rep(NA, nrow(df)))
  }
}

# Display key clinical variables
clinical_summary <- data.frame(
  Patient = meta_sub$patientID,
  Age = get_col_safe(meta_sub, "years_to_birth") * -1,
  Race = get_col_safe(meta_sub, "race"),
  Stage = get_col_safe(meta_sub, "pathology_T_stage"),
  Gleason_Score = get_col_safe(meta_sub, "gleason_score"),
  PSA = get_col_safe(meta_sub, "PSA"),
  Vital_Status = get_col_safe(meta_sub, "vital_status"),
  Days_to_Death = get_col_safe(meta_sub, "days_to_death"),
  Days_to_Followup = get_col_safe(meta_sub, "days_to_last_followup"),
  ERG_status = get_col_safe(meta_sub, "ERG_status")
)

kable(clinical_summary, 
      caption = "Clinical Characteristics of Analyzed Samples",
      align = "lcccccccc") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 10)
```

### Clinical Context: Prostate Cancer Cohort

These samples represent **prostate adenocarcinoma (PRAD)** cases from TCGA:

**Key Clinical Variables**:

- **Gleason Score**: Histological grading system (6-10)
  - 6: Well-differentiated, low grade
  - 7: Intermediate grade
  - 8-10: Poorly differentiated, high grade
  - Higher scores correlate with aggressive disease

- **Pathology T Stage**: Extent of primary tumor
  - T2: Confined to prostate
  - T3: Extends beyond prostate capsule
  - T4: Invades adjacent structures

- **PSA**: Prostate-specific antigen (tumor marker)
  - Normal: <4 ng/mL
  - Elevated: >10 ng/mL suggests higher tumor burden

### Linking Spatial Features to Clinical Outcomes

```{r clinical-spatial-integration}
# Get available clinical columns
available_cols <- colnames(meta_sub)
clinical_cols <- c("patientID", "gleason_score", "pathology_T_stage", "PSA", 
                   "vital_status", "days_to_death", "days_to_last_followup", 
                   "years_to_birth", "race", "ERG_status")
cols_to_select <- intersect(clinical_cols, available_cols)

# Add clinical data to multiscale features
multiscale_clinical <- multiscale_features %>%
  mutate(patientID = sub("^(TCGA-[0-9A-Z]{2}-[0-9A-Z]{4}).*", "\\1", slide_id)) %>%
  left_join(
    meta_sub %>% 
      as.data.frame() %>%
      select(all_of(cols_to_select)),
    by = "patientID"
  )

# Add derived categorical variables if base columns exist
if ("gleason_score" %in% colnames(multiscale_clinical)) {
  multiscale_clinical <- multiscale_clinical %>%
    mutate(
      Gleason_Category = case_when(
        is.na(gleason_score) ~ "Unknown",
        gleason_score <= 6 ~ "Low (6)",
        gleason_score == 7 ~ "Intermediate (7)",
        gleason_score >= 8 ~ "High (8-10)",
        TRUE ~ "Unknown"
      )
    )
}

if ("pathology_T_stage" %in% colnames(multiscale_clinical)) {
  multiscale_clinical <- multiscale_clinical %>%
    mutate(
      T_Stage_Group = case_when(
        is.na(pathology_T_stage) ~ "Unknown",
        grepl("T2", pathology_T_stage) ~ "T2 (Confined)",
        grepl("T3", pathology_T_stage) ~ "T3 (Extraprostatic)",
        grepl("T4", pathology_T_stage) ~ "T4 (Invasive)",
        TRUE ~ "Unknown"
      )
    )
}

if ("PSA" %in% colnames(multiscale_clinical)) {
  multiscale_clinical <- multiscale_clinical %>%
    mutate(
      PSA_Category = case_when(
        is.na(PSA) ~ "Unknown",
        PSA < 4 ~ "Normal (<4)",
        PSA >= 4 & PSA < 10 ~ "Elevated (4-10)",
        PSA >= 10 ~ "High (≥10)",
        TRUE ~ "Unknown"
      )
    )
}

if ("years_to_birth" %in% colnames(multiscale_clinical)) {
  multiscale_clinical <- multiscale_clinical %>%
    mutate(Age = years_to_birth * -1)
}

# Display integration summary
n_with_gleason <- sum(!is.na(multiscale_clinical$gleason_score) & 
                      multiscale_clinical$gleason_score != "Unknown")
cat("Successfully integrated clinical data for", nrow(multiscale_clinical), "samples\n")
if ("gleason_score" %in% colnames(multiscale_clinical)) {
  cat("Samples with Gleason score:", n_with_gleason, "\n")
}
```



## Clinical-Spatial Correlation Analysis

### Spatial Features by Gleason Score

```{r spatial-by-gleason, fig.width=12, fig.height=8, eval="Gleason_Category" %in% colnames(multiscale_clinical)}
# Check if we have Gleason score data
if ("Gleason_Category" %in% colnames(multiscale_clinical)) {
  # Extract key spatial features for comparison
  spatial_summary <- subset(multiscale_clinical, 
       !is.na(gleason_score) & Gleason_Category != "Unknown") %>%
    select(patientID, Gleason_Category, 
           matches("mean_deg_inflammatory_r100"),
           matches("mean_deg_neoplastic_r100"),
           matches("mean_mix_inflammatory_r100"),
           matches("mean_mix_neoplastic_r100")) %>%
    pivot_longer(cols = -c(patientID, Gleason_Category),
                 names_to = "Feature",
                 values_to = "Value") %>%
    mutate(
      Feature_Label = case_when(
        grepl("deg_inflammatory", Feature) ~ "Inflammatory Degree (r100)",
        grepl("deg_neoplastic", Feature) ~ "Neoplastic Degree (r100)",
        grepl("mix_inflammatory", Feature) ~ "Inflammatory Mixing (r100)",
        grepl("mix_neoplastic", Feature) ~ "Neoplastic Mixing (r100)",
        TRUE ~ Feature
      )
    )
  
  if (nrow(spatial_summary) > 0) {
    ggplot(spatial_summary, aes(x = Gleason_Category, y = Value, fill = Gleason_Category)) +
      geom_boxplot() +
      geom_point(position = position_jitter(width = 0.2), alpha = 0.6) +
      facet_wrap(~ Feature_Label, scales = "free_y", ncol = 2) +
      scale_fill_brewer(palette = "YlOrRd") +
      theme_bw() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1),
            legend.position = "none") +
      labs(title = "Spatial Features by Gleason Score",
           subtitle = "100 μm radius features",
           x = "Gleason Score Category",
           y = "Feature Value")
  } else {
    cat("No Gleason score data available for spatial analysis\n")
  }
} else {
  cat("Gleason score not available in clinical data\n")
}
```

### Interpretation: Spatial Organization and Tumor Grade

**Clinical Significance of Gleason Score**:

- **Low Grade (≤6)**: Well-differentiated, gland-forming architecture
- **Intermediate (7)**: Mixed differentiation, common grade
- **High Grade (≥8)**: Poorly differentiated, loss of glandular structure

**Expected Spatial Patterns by Grade**:

1. **Inflammatory Features**:
   - **Higher grade**: May show increased immune infiltration (response to more aggressive tumor)
   - **Lower grade**: May have less immune response
   - **Mixing**: Higher in aggressive tumors (breakdown of tissue barriers)

2. **Neoplastic Features**:
   - **Higher grade**: More dispersed distribution (infiltrative growth)
   - **Lower grade**: Organized tumor nests (glandular architecture)
   - **Degree patterns**: Reflect growth pattern differences

**Observed Patterns**:

```{r gleason-observations, eval=exists("spatial_summary") && nrow(spatial_summary) > 0}
# Calculate mean differences if spatial_summary exists
if (exists("spatial_summary") && nrow(spatial_summary) > 0) {
  gleason_means <- spatial_summary %>%
    group_by(Gleason_Category, Feature_Label) %>%
    summarize(Mean_Value = mean(Value, na.rm = TRUE), .groups = "drop") %>%
    pivot_wider(names_from = Gleason_Category, values_from = Mean_Value)
  
  kable(gleason_means,
        caption = "Mean Spatial Features by Gleason Grade",
        digits = 2) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = FALSE)
} else {
  cat("Spatial summary data not available for Gleason analysis\n")
}
```

**Clinical Implications**:

- Spatial features may complement Gleason scoring
- Quantitative metrics could improve risk stratification
- Differences suggest grade-specific microenvironmental phenotypes

### PCA with Clinical Annotation

```{r pca-clinical, fig.width=12, fig.height=5}
# Perform PCA with clinical annotation
# Select only columns that exist in multiscale_clinical
clinical_cols_available <- c("slide_id", "Gleason_Category", "T_Stage_Group", "PSA_Category", "Age", "ERG_status")
clinical_cols_present <- intersect(clinical_cols_available, colnames(multiscale_clinical))

pca_scores_clinical <- as.data.frame(pca_result$x) %>%
  mutate(slide_id = multiscale_features$slide_id) %>%
  left_join(
    multiscale_clinical %>% select(all_of(clinical_cols_present)),
    by = "slide_id"
  )

# Create plots based on available clinical variables
has_gleason <- "Gleason_Category" %in% colnames(pca_scores_clinical)
has_tstage <- "T_Stage_Group" %in% colnames(pca_scores_clinical)
has_psa <- "PSA_Category" %in% colnames(pca_scores_clinical)
has_age <- "Age" %in% colnames(pca_scores_clinical)

# Plot 1: Gleason and T Stage (if available)
if (has_gleason && has_tstage) {
  p1 <- ggplot(pca_scores_clinical, aes(x = PC1, y = PC2, color = Gleason_Category, shape = T_Stage_Group)) +
    geom_point(size = 4, alpha = 0.8) +
    scale_color_brewer(palette = "YlOrRd") +
    theme_bw() +
    labs(title = "PCA colored by Gleason Score and T Stage",
         x = paste0("PC1 (", round(var_explained[1] * 100, 1), "%)"),
         y = paste0("PC2 (", round(var_explained[2] * 100, 1), "%)"),
         color = "Gleason Grade",
         shape = "T Stage")
} else if (has_gleason) {
  p1 <- ggplot(pca_scores_clinical, aes(x = PC1, y = PC2, color = Gleason_Category)) +
    geom_point(size = 4, alpha = 0.8) +
    scale_color_brewer(palette = "YlOrRd") +
    theme_bw() +
    labs(title = "PCA colored by Gleason Score",
         x = paste0("PC1 (", round(var_explained[1] * 100, 1), "%)"),
         y = paste0("PC2 (", round(var_explained[2] * 100, 1), "%)"),
         color = "Gleason Grade")
} else {
  p1 <- ggplot(pca_scores_clinical, aes(x = PC1, y = PC2)) +
    geom_point(size = 4, alpha = 0.8, color = "steelblue") +
    theme_bw() +
    labs(title = "PCA Scores",
         x = paste0("PC1 (", round(var_explained[1] * 100, 1), "%)"),
         y = paste0("PC2 (", round(var_explained[2] * 100, 1), "%)"))
}

# Plot 2: PSA and Age (if available)
if (has_psa && has_age) {
  p2 <- ggplot(pca_scores_clinical, aes(x = PC1, y = PC2, color = PSA_Category, size = Age)) +
    geom_point(alpha = 0.8) +
    scale_color_brewer(palette = "RdYlBu") +
    theme_bw() +
    labs(title = "PCA colored by PSA Level and Age",
         x = paste0("PC1 (", round(var_explained[1] * 100, 1), "%)"),
         y = paste0("PC2 (", round(var_explained[2] * 100, 1), "%)"),
         color = "PSA Level",
         size = "Age")
  p1 + p2
} else if (has_age) {
  p2 <- ggplot(pca_scores_clinical, aes(x = PC1, y = PC2, size = Age)) +
    geom_point(alpha = 0.8, color = "steelblue") +
    theme_bw() +
    labs(title = "PCA colored by Age",
         x = paste0("PC1 (", round(var_explained[1] * 100, 1), "%)"),
         y = paste0("PC2 (", round(var_explained[2] * 100, 1), "%)"),
         size = "Age")
  p1 + p2
} else {
  print(p1)
  cat("\nNote: PSA and/or Age data not available in clinical metadata\n")
}
```

### Interpretation: Clinical Correlates of Spatial Phenotypes

**Key Questions**:

1. **Do samples cluster by clinical characteristics?**
   - Samples with similar Gleason scores cluster together → spatial features track grade
   - Random distribution → spatial heterogeneity independent of grade

2. **Does PC1 or PC2 separate clinical groups?**
   - Separation along PC1 → primary axis of spatial variation relates to clinical features
   - No clear separation → spatial features provide complementary information

3. **Are there grade-specific spatial phenotypes?**
   - High grade outliers → unique aggressive TME architectures
   - Low grade clustering → conserved well-differentiated patterns

**Clinical Value**:

- **Prognostic Refinement**: Spatial features may substratify within Gleason grades
- **Treatment Selection**: TME architecture could predict therapy response
- **Biomarker Development**: Combining grade + spatial metrics improves prediction

### Immune Infiltration by Clinical Characteristics

```{r immune-clinical, fig.width=12, fig.height=4, eval="Gleason_Category" %in% colnames(multiscale_clinical) && sum(!is.na(multiscale_clinical$gleason_score)) > 0}
# Analyze immune features by clinical variables only if data available
if ("Gleason_Category" %in% colnames(multiscale_clinical) && 
    sum(!is.na(multiscale_clinical$gleason_score)) > 0) {
  
  immune_features <- multiscale_clinical[!is.na(multiscale_clinical$gleason_score) & 
                    multiscale_clinical$Gleason_Category != "Unknown", ] %>%
    select(patientID, Gleason_Category, T_Stage_Group,
           matches("mean_deg_inflammatory_r\\d+")) %>%
    pivot_longer(cols = matches("mean_deg"),
                 names_to = "Radius",
                 values_to = "Inflammatory_Degree") %>%
    mutate(
      Radius_Value = as.numeric(gsub(".*_r(\\d+)$", "\\1", Radius))
    )
  
  p1 <- ggplot(immune_features, 
         aes(x = factor(Radius_Value), y = Inflammatory_Degree, 
             color = Gleason_Category, group = Gleason_Category)) +
    geom_line(stat = "summary", fun = "mean", size = 1.5) +
    geom_point(stat = "summary", fun = "mean", size = 3) +
    scale_color_brewer(palette = "YlOrRd") +
    theme_bw() +
    labs(title = "Immune Infiltration Across Scales by Grade",
         x = "Radius (μm)",
         y = "Mean Inflammatory Degree",
         color = "Gleason Grade")
  
  if ("T_Stage_Group" %in% colnames(immune_features) && 
      sum(!is.na(immune_features$T_Stage_Group) & 
          immune_features$T_Stage_Group != "Unknown") > 0) {
    p2 <- ggplot(immune_features %>% filter(!is.na(T_Stage_Group) & T_Stage_Group != "Unknown"), 
           aes(x = factor(Radius_Value), y = Inflammatory_Degree, 
               color = T_Stage_Group, group = T_Stage_Group)) +
      geom_line(stat = "summary", fun = "mean", size = 1.5) +
      geom_point(stat = "summary", fun = "mean", size = 3) +
      scale_color_brewer(palette = "Set1") +
      theme_bw() +
      labs(title = "Immune Infiltration Across Scales by T Stage",
           x = "Radius (μm)",
           y = "Mean Inflammatory Degree",
           color = "T Stage")
    p1 + p2
  } else {
    print(p1)
  }
} else {
  cat("Clinical data (Gleason score) not available for immune infiltration analysis\n")
}
```

### Interpretation: Immune Infiltration and Disease Characteristics

**Clinical Context**:

In prostate cancer:
- Immune infiltration patterns vary with disease progression
- Higher grade tumors may show different immune microenvironments
- T stage reflects local invasion extent

**Potential Patterns**:

1. **Grade-dependent immune response**:
   - Higher Gleason → may show increased or decreased immune infiltration
   - Pattern reflects tumor immunogenicity and immune evasion

2. **Stage-dependent patterns**:
   - T3/T4 (invasive) → altered immune landscape at invasive front
   - T2 (confined) → more organized immune response

3. **Scale-dependent differences**:
   - Small scales (20-40 μm): Direct immune-tumor interactions
   - Large scales (80-100 μm): Tissue-level immune organization

**Therapeutic Implications**:

- **High immune infiltration + high mixing**: May respond to immunotherapy
- **Low immune infiltration**: Consider immune-activating strategies
- **Compartmentalized immunity**: Potential barriers to treatment

### Degree-Mixing Relationship by Clinical Group

```{r degree-mixing-clinical, fig.width=12, fig.height=8, eval="Gleason_Category" %in% colnames(multiscale_clinical) && exists("combined")}
# Degree-mixing by Gleason score if data available
if ("Gleason_Category" %in% colnames(multiscale_clinical) && exists("combined")) {
  degree_mixing_clinical <- combined %>%
    left_join(
      multiscale_clinical %>% select(slide_id, any_of(c("Gleason_Category", "T_Stage_Group"))),
      by = "slide_id"
    ) %>%
      .[!is.na(.$Gleason_Category) & .$Gleason_Category != "Unknown", ]
  
  if (nrow(degree_mixing_clinical) > 0) {
    ggplot(degree_mixing_clinical, 
           aes(x = mean_degree, y = mean_mixing, color = Gleason_Category)) +
      geom_point(size = 3, alpha = 0.7) +
      geom_smooth(method = "lm", se = TRUE, alpha = 0.2) +
      facet_wrap(~ cell_type, scales = "free") +
      scale_color_brewer(palette = "YlOrRd") +
      theme_bw() +
      labs(title = "Degree-Mixing Relationship by Cell Type and Gleason Score",
           x = "Mean Degree (across scales)",
           y = "Mean Mixing Index (across scales)",
           color = "Gleason Grade")
  } else {
    cat("No data available for degree-mixing analysis by Gleason score\n")
  }
} else {
  cat("Clinical data or degree-mixing data not available\n")
}
```

### Interpretation: Degree-Mixing Relationship

1. **Positive Correlation** (degree ↑, mixing ↑):
   - More neighbors = more opportunity for diverse neighbors
   - Dense regions tend to be more heterogeneous
   - Common in well-mixed tissues

2. **Negative Correlation** (degree ↑, mixing ↓):
   - Dense regions are homogeneous (single cell type clusters)
   - Sparse regions have diverse neighbors
   - Suggests strong cell type clustering

3. **No Correlation**:
   - Density and diversity are independent
   - Complex tissue organization

**Cell Type-Specific Patterns**:

**Inflammatory Cells**:
- Often positive correlation
- When immune infiltration increases (higher degree), more mixing occurs
- Reflects infiltration into other compartments vs. isolated clusters

**Stromal Cells**:
- Usually weak or no correlation
- Ubiquitous distribution regardless of density
- Consistent mixing across tissue regions

**Neoplastic Cells**:
- Variable patterns, diagnostically important:
  - **Positive**: Infiltrative growth (more tumor cells → more mixing with stroma)
  - **Negative**: Expansive growth (dense tumor nests, homogeneous)
  - Pattern correlates with prognosis in many cancers

**Clinical Implications**:

- **Strong correlations**: Predictable spatial organization
  - Easier to target therapeutically
  - Uniform treatment response expected

- **Weak correlations**: Complex, heterogeneous organization
  - Multiple spatial phenotypes coexist
  - May require combination approaches

- **Sample-specific patterns**: 
  - Different degree-mixing relationships across samples
  - Reflects distinct TME architectures
  - Could define patient subgroups

**Research Questions**:

1. Do different relationships predict clinical outcomes?
2. Does the relationship change with treatment?
3. Are there cell type-specific thresholds (e.g., inflammatory degree >10 → high mixing)?
4. Can we use the relationship to identify therapeutic vulnerabilities?

# Summary and Recommendations

## Key Findings

This vignette demonstrated comprehensive exploratory data analysis of 
graph-based spatial features from histopathology images integrated with 
clinical metadata. Key findings include:

1. **Graph Structure**: Analysis of cell-cell connectivity through degree 
   and mixing indices reveals tissue organization principles
2. **Cell Type Composition**: Distribution of different cell types and their 
   spatial interactions define TME phenotypes
3. **Multiscale Features**: How graph properties change across different 
   spatial scales (20-100 μm) captures hierarchical organization
4. **Dimensionality Reduction**: PCA reveals major sources of variation 
   across samples and identifies spatial phenotypes
5. **Feature Relationships**: Correlation and interaction patterns between 
   different graph metrics suggest biological processes
6. **Clinical Integration**: Spatial features show associations with Gleason 
   grade, T stage, and other clinical characteristics
7. **Prognostic Potential**: Combined spatial-clinical features may improve 
   risk stratification beyond conventional grading

## Applications

These graph-based features can be used for:

### Clinical Applications
- **Enhanced Risk Stratification**: Combine Gleason grade with spatial features
  - Identify high-risk patients within intermediate Gleason scores
  - Refine prognosis using TME architecture
- **Treatment Selection**: Match spatial features to therapy mechanisms
  - High immune infiltration + high mixing → immunotherapy candidates
  - Dense stroma + low mixing → stromal targeting + chemotherapy
  - Grade-specific spatial patterns → tailored treatment approaches
- **Response Prediction**: Spatial features predict treatment outcomes
  - Immune infiltration patterns → immunotherapy response
  - Tumor-stroma organization → chemotherapy penetration
- **Monitoring**: Changes in spatial organization track disease progression
  - Serial biopsies showing spatial evolution
  - Treatment-induced TME remodeling
- **Clinical Trial Design**: 
  - Spatial phenotypes as inclusion criteria
  - TME-based patient matching to therapy mechanisms
  - Spatial biomarkers as surrogate endpoints

### Research Applications
- **Tumor Microenvironment Characterization**: Quantify TME architecture
- **Spatial Heterogeneity Quantification**: Measure intra-tumor diversity
- **Predictive Modeling**: Machine learning on spatial features
- **Cohort Comparison**: Standardized metrics across studies
- **Mechanism Discovery**: Link spatial patterns to molecular pathways

## Best Practices

### Feature Selection
1. **Start with domain knowledge**: Choose features with biological rationale
2. **Remove redundancy**: Use correlation analysis to reduce feature set
3. **Consider multiple scales**: Different scales capture different biology
4. **Validate consistently**: Test in independent cohorts

### Interpretation Guidelines
1. **Context matters**: Same feature means different things in different tissues
2. **Multiple features**: No single metric captures complete picture
3. **Quality control**: Technical artifacts can masquerade as biology
4. **Statistical power**: Ensure adequate sample size for robust conclusions

### Common Pitfalls to Avoid
1. **Over-interpreting single samples**: Need population-level validation
2. **Ignoring technical variation**: Batch effects, staining variation
3. **Assuming causality**: Correlations don't imply mechanisms
4. **Feature fishing**: Multiple testing correction for many features

## Future Directions

### Methodological Advances
- **3D imaging**: Capture true spatial relationships (not just 2D sections)
- **Temporal analysis**: Longitudinal studies of spatial evolution
- **Multi-modal integration**: Combine spatial + molecular data
- **Deep learning**: Automated feature discovery and interpretation

### Clinical Translation
- **Biomarker Validation**: Prospective studies linking features to outcomes
- **Clinical Trials**: Spatial features as inclusion criteria or endpoints
- **Diagnostic Tools**: Integration into pathology workflows
- **Personalized Medicine**: Spatial phenotype-guided treatment decisions

## Recommended Next Steps

After completing this exploratory analysis:

1. **Hypothesis Generation**: 
   - Which features distinguish your samples?
   - What biological processes might explain the patterns?

2. **Statistical Testing**:
   - Formal tests for differences between groups
   - Correct for multiple comparisons

3. **Clinical Correlation**:
   - Link spatial features to patient outcomes
   - Survival analysis, response rates

4. **Mechanistic Studies**:
   - Validate in experimental models
   - Manipulate spatial organization, measure effects

5. **Independent Validation**:
   - Test findings in separate cohort
   - Confirm reproducibility

## Resources for Further Learning

- **Graph Theory in Biology**: Network analysis fundamentals
- **Spatial Statistics**: Methods for spatial data analysis
- **Computational Pathology**: Digital pathology and AI
- **Tumor Immunology**: TME biology and immune interactions
- **Clinical Trial Design**: Translation of biomarkers to practice

# Session Information

```{r session-info}
sessionInfo()
```
