---
title: "Multi-modal Analysis: Cell-Graph Features with Genomic and Clinical Data"
author: "Sehyun Oh"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteIndexEntry{Multi-modal Graph Feature Analysis}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
output:
  BiocStyle::html_document:
    number_sections: yes
    toc: yes
    toc_depth: 4
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    comment   = "#>",
    collapse  = TRUE,
    message   = FALSE,
    warning   = FALSE,
    fig.width = 10,
    fig.height = 7
)
```

# Overview

This vignette demonstrates how to import cell-graph topology and
slide-level graph features into Bioconductor objects and integrate them with 
genomic and clinical data for downstream statistical analysis.

**Data sources used:**

| Layer | File(s) | R object |
|-------|---------|----------|
| Cell-graph topology | This package (`inst/extdata/*_graph.h5`) | `igraph` per slide |
| Patient-level graph features | This package (`inst/extdata`) | `SummarizedExperiment` |
| Genomic data (RNA-seq) | curatedTCGAData | `MultiAssayExperiment` |
| Clinical metadata | curatedTCGAData `colData` | `DataFrame` |

# Setup

```{r packages}
suppressPackageStartupMessages({
    
    ## Bioconductor
    library(SummarizedExperiment)
    library(MultiAssayExperiment)
    library(S4Vectors)

    ## Graph analysis
    library(igraph)

    ## Genomic data
    library(curatedTCGAData)
    library(TCGAutils)

    ## Visualization & stats
    library(ggplot2)
    library(ComplexHeatmap)
    library(survival)
    library(survminer)
    library(dplyr)
})
```

## File paths

```{r paths}
dataDir <- system.file("extdata", package = "CellGraphR")
h5_files <- list.files(dataDir, pattern = "_graph\\.h5$", full.names = TRUE)
csv_file <- file.path(dataDir,
                      "all_fullcohort_combined_expanded_patient_level_features_2.csv")

cat("H5 graph files found:", length(h5_files), "\n")
basename(h5_files)
```

# Import Graph Data

## Cell-graph topology (`igraph`)

Each `.h5` file stores the full spatial cell graph for one whole-slide
image. `importAllGraphH5()` reads them in bulk and returns a named list
of `igraph` objects — one per patient.

```{r import-topology}
graphs <- importAllGraphH5(h5_files, as_igraph = TRUE, verbose = TRUE)
```

Inspect a single slide's graph:

```{r topology-summary}
g <- graphs[[1]]

cat("Patient:  ", igraph::graph_attr(g, "patient_id"), "\n")
cat("Nodes:    ", igraph::vcount(g), "(cells)\n")
cat("Edges:    ", igraph::ecount(g), "(cell-cell contacts)\n")
cat("Density:  ", round(igraph::edge_density(g), 5), "\n")

## Degree distribution
## The degree is the total number of edges incident to a vertex.
deg <- igraph::degree(g)
summary(deg)
```

```{r degree-dist-plot}
ggplot(data.frame(degree = deg), aes(x = degree)) +
    geom_histogram(bins = 40, fill = "#2C7BB6", color = "white") +
    labs(title = paste("Degree distribution —",
                       igraph::graph_attr(g, "patient_id")),
         x = "Node degree (# neighbouring cells)",
         y = "Count") +
    theme_bw()
```

### Per-slide graph-level summaries

Compute summary statistics across all imported graphs:

```{r graph-summary-table}
graph_summary <- do.call(rbind, lapply(graphs, function(gi) {
    data.frame(
        patient_id       = igraph::graph_attr(gi, "patient_id"),
        n_cells          = igraph::vcount(gi),
        n_edges          = igraph::ecount(gi),
        edge_density     = igraph::edge_density(gi),
        mean_degree      = mean(igraph::degree(gi)),
        transitivity     = igraph::transitivity(gi, type = "global"),
        n_components     = igraph::components(gi)$no,
        stringsAsFactors = FALSE
    )
}))
rownames(graph_summary) <- graph_summary$patient_id
print(graph_summary)
```

## Patient-level graph features (`SummarizedExperiment`)

`importGraphFeatures()` reads the pre-computed CSV summary and returns a
`SummarizedExperiment` with features as rows and patients as columns.
Features are automatically annotated by thematic group in `rowData`.

```{r import-features}
graph_se <- importGraphFeatures(csv_file, feature_groups = TRUE)
graph_se
```

```{r rowdata-preview}
## Feature groups
table(rowData(graph_se)$feature_group)
```

### Feature matrix overview

```{r feature-heatmap, fig.height=10, eval=FALSE}
mat_raw    <- as.matrix(assay(graph_se, "graph_features"))
mat_scaled <- base::t(base::scale(base::t(mat_raw)))

## Drop rows/columns that are all-NA after scaling (zero-variance or all-NA features)
keep_rows <- rowSums(is.finite(mat_scaled)) > 0
keep_cols <- colSums(is.finite(mat_scaled)) > 0
mat_scaled <- mat_scaled[keep_rows, keep_cols, drop = FALSE]

## Color-code columns by feature group for annotation
row_anno <- rowAnnotation(
    group = rowData(graph_se)$feature_group[keep_rows],
    col   = list(group = setNames(
        rainbow(length(unique(rowData(graph_se)$feature_group))),
        unique(rowData(graph_se)$feature_group)))
)

Heatmap(mat_scaled,
        name            = "z-score",
        right_annotation = row_anno,
        show_row_names  = FALSE,
        show_column_names = TRUE,
        column_title    = "Patients",
        row_title       = paste(sum(keep_rows), "graph features"),
        cluster_rows    = TRUE,
        cluster_columns = TRUE)
```


# Multi-modal Integration

## Load TCGA genomic + clinical data

```{r load-tcga, eval=FALSE}
## Download TCGA-PRAD RNA-seq and clinical data
mae_prad <- curatedTCGAData::curatedTCGAData(
    diseaseCode = "PRAD",
    assays      = "RNASeq2GeneNorm*",
    version     = "2.0.1",
    dry.run     = FALSE
)
```

```{r}
mae <- mae_prad[, colnames(graph_se),]

prad_pids <- colnames(graph_se)
clin <- colData(rna_se)
rna_mat <- assay(rna_se[[1]])
```

```{r load-tcga-show, echo=FALSE, eval=TRUE}
## For illustration, create a minimal placeholder MAE
set.seed(42)
n_genes    <- 200
prad_pids  <- colData(graph_se)$patient_id
rna_mat    <- matrix(rpois(n_genes * length(prad_pids), lambda = 50),
                     nrow = n_genes,
                     dimnames = list(paste0("GENE", seq_len(n_genes)),
                                     prad_pids))
rna_se     <- SummarizedExperiment(assays = list(counts = rna_mat),
                                   colData = DataFrame(patient_id = prad_pids,
                                                       row.names = prad_pids))

## Simulated clinical metadata
clin <- DataFrame(
    patient_id    = prad_pids,
    gleason_score = sample(6:10, length(prad_pids), replace = TRUE),
    pathologic_T  = sample(c("T2", "T3", "T4"), length(prad_pids), replace = TRUE),
    days_to_death       = rexp(length(prad_pids), rate = 1/60),
    vital_status     = rbinom(length(prad_pids), 1, 0.3),
    row.names     = prad_pids
)

smap_rna <- DataFrame(assay   = "RNASeq2GeneNorm",
                      primary = prad_pids,
                      colname = prad_pids)

mae_prad <- MultiAssayExperiment(
    experiments = ExperimentList(RNASeq2GeneNorm = rna_se),
    colData     = clin,
    sampleMap   = smap_rna
)
cat("MAE created for illustration:\n")
mae
```

## Add graph features to the MAE

`addGraphToMAE()` aligns patients by ID, subsets to common samples, and
inserts the `SummarizedExperiment` as a new experiment slot.

```{r add-graph}
mae_multi <- addGraphToMAE(mae, graph_se,
                           experiment_name = "graph_features")
mae_multi
```

# Downstream Statistical Analysis
## Dimensionality reduction on graph features
### PCA

```{r pca}
feat_mat <- assay(mae_multi[["graph_features"]])

## Remove zero-variance features (common with indicator columns)
rv      <- matrixStats::rowVars(feat_mat, na.rm = TRUE)
feat_nz <- feat_mat[rv > 1e-8, ]
feat_nz <- na.omit(feat_nz)

pca_res <- prcomp(t(feat_nz), scale. = TRUE, center = TRUE)

pca_df <- data.frame(
    PC1        = pca_res$x[, 1],
    PC2        = pca_res$x[, 2],
    patient_id = rownames(pca_res$x)
)
pca_df <- merge(pca_df,
                as.data.frame(colData(mae_multi)),
                by.x = "patient_id", by.y = "row.names",
                all.x = TRUE)

var_exp <- round(100 * pca_res$sdev^2 / sum(pca_res$sdev^2), 1)

ggplot(pca_df, aes(x = PC1, y = PC2, color = factor(gleason_score))) +
    geom_point(size = 3, alpha = 0.8) +
    scale_color_brewer(palette = "RdYlBu", name = "Gleason score") +
    labs(title = "PCA of patient-level cell-graph features",
         x = paste0("PC1 (", var_exp[1], "%)"),
         y = paste0("PC2 (", var_exp[2], "%)")) +
    theme_bw()
```

```{r pca-scree}
scree_df <- data.frame(
    PC          = seq_along(var_exp)[seq_len(min(15, length(var_exp)))],
    var_explained = var_exp[seq_len(min(15, length(var_exp)))]
)
ggplot(scree_df, aes(x = PC, y = var_explained)) +
    geom_col(fill = "#4DAF4A") +
    geom_line(aes(group = 1), color = "grey30") +
    geom_point() +
    labs(title = "Scree plot — graph features PCA",
         x = "Principal component", y = "Variance explained (%)") +
    theme_bw()
```

## Association with clinical variables

### Kruskal-Wallis test: graph features ~ Gleason score

```{r kw-test}
clin_df     <- as.data.frame(colData(mae_multi))
shared_pids <- intersect(colnames(feat_mat), rownames(clin_df))

feat_sub  <- feat_mat[, shared_pids, drop = FALSE]
clin_sub  <- clin_df[shared_pids, , drop = FALSE]

## Test each feature
kw_pvals <- apply(feat_sub, 1, function(x) {
    tryCatch(
        kruskal.test(x ~ factor(clin_sub$gleason_score))$p.value,
        error = function(e) NA_real_
    )
})

kw_df <- data.frame(
    feature       = names(kw_pvals),
    pvalue        = kw_pvals,
    padj          = p.adjust(kw_pvals, method = "BH"),
    feature_group = rowData(graph_se)[names(kw_pvals), "feature_group"]
)
kw_df <- kw_df[order(kw_df$padj), ]

cat("Features significantly associated with Gleason score (FDR < 0.05):\n")
sig_features <- kw_df[!is.na(kw_df$padj) & kw_df$padj < 0.05, ]
print(sig_features)
```

```{r volcano-gleason}
kw_df$neg_log10_padj <- -log10(kw_df$padj + 1e-10)

ggplot(kw_df, aes(x = neg_log10_padj,
                  y = reorder(feature, neg_log10_padj),
                  color = feature_group)) +
    geom_point(size = 2) +
    geom_vline(xintercept = -log10(0.05), linetype = "dashed",
               color = "red", linewidth = 0.7) +
    labs(title = "Graph features associated with Gleason score",
         x = "-log10(adjusted p-value)",
         y = NULL, color = "Feature group") +
    theme_bw() +
    theme(axis.text.y = element_text(size = 7))
```

### Boxplots for top features

```{r top-feature-boxplot, fig.height=5}
# Fall back to top features by raw p-value if none survive FDR correction
top_feats <- if (nrow(sig_features) > 0) {
    head(sig_features$feature, 4)
} else {
    message("No features pass FDR < 0.05; showing top 4 by raw p-value instead.")
    head(kw_df$feature, 4)
}

plot_df <- data.frame(
    t(feat_sub[top_feats, , drop = FALSE]),
    gleason = factor(clin_sub$gleason_score),
    check.names = FALSE
)

tidyr::pivot_longer(plot_df,
                    cols      = all_of(top_feats),
                    names_to  = "feature",
                    values_to = "value") |>
    ggplot(aes(x = gleason, y = value, fill = gleason)) +
    geom_boxplot(outlier.size = 1, alpha = 0.7) +
    facet_wrap(~feature, scales = "free_y", ncol = 2) +
    scale_fill_brewer(palette = "Set2") +
    labs(title = "Top graph features by Gleason score",
         x = "Gleason score", y = "Feature value") +
    theme_bw() + theme(legend.position = "none")
```

## Survival analysis

### Cox proportional hazards with graph-feature PCs

```{r survival-cox}
pca_pids <- intersect(shared_pids, rownames(pca_res$x))
n_pcs    <- min(5L, ncol(pca_res$x))

surv_df <- data.frame(
    days_to_death   = clin_sub[pca_pids, "days_to_death"],
    vital_status = clin_sub[pca_pids, "vital_status"],
    pca_res$x[pca_pids, seq_len(n_pcs), drop = FALSE]
)
surv_df <- surv_df[complete.cases(surv_df), ]

cox_fit <- coxph(Surv(days_to_death, vital_status) ~ .,
                 data = surv_df)
summary(cox_fit)
```

### Log-rank test: PC1 high vs. low

```{r km-plot}
med_pc1 <- median(surv_df$PC1)
surv_df$pc1_group <- ifelse(surv_df$PC1 > med_pc1, "High", "Low")

km_fit <- survfit(Surv(days_to_death, vital_status) ~ pc1_group, data = surv_df)

ggsurvplot(km_fit,
           data       = surv_df,
           pval       = TRUE,
           risk.table = TRUE,
           palette    = c("#E41A1C", "#377EB8"),
           title      = "Overall survival by graph-feature PC1",
           xlab       = "Time (months)",
           legend.labs = c("PC1 High", "PC1 Low"))
```

## Correlation with RNA-seq

Correlate each graph feature with the first RNA-seq PC (as a proxy for
transcriptomic variation).

```{r rna-correlation}
rna_mat_sub <- assay(mae_multi[["RNASeq2GeneNorm"]])[, shared_pids]
rna_log     <- log1p(rna_mat_sub)
rna_log     <- rna_log[apply(rna_log, 1, var) > 0, ]   # drop zero-variance genes
rna_pc1     <- prcomp(t(rna_log), scale. = TRUE)$x[, 1]

feat_rna_cor <- apply(feat_sub, 1, function(x) {
    tryCatch(
        cor(x, rna_pc1, use = "complete.obs", method = "spearman"),
        error = function(e) NA_real_
    )
})

cor_df <- data.frame(
    feature       = names(feat_rna_cor),
    spearman_rho  = feat_rna_cor,
    feature_group = rowData(graph_se)[names(feat_rna_cor), "feature_group"]
)
cor_df <- cor_df[order(abs(cor_df$spearman_rho), decreasing = TRUE), ]

cat("Top features correlated with RNA-seq PC1:\n")
print(head(cor_df, 10))
```

```{r rna-cor-plot}
ggplot(cor_df, aes(x = spearman_rho, fill = feature_group)) +
    geom_histogram(bins = 30, color = "white") +
    geom_vline(xintercept = 0, linetype = "dashed") +
    labs(title = "Distribution of graph-feature ~ RNA-seq PC1 correlations",
         x = "Spearman ρ", y = "Count", fill = "Feature group") +
    theme_bw()
```

# Session Info

```{r session-info}
sessionInfo()
```
