---
title: "Analyzing Cell Graphs with igraph"
author: "Sehyun Oh"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteIndexEntry{Analyzing Cell Graphs with igraph}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
output:
  BiocStyle::html_document:
    number_sections: yes
    toc: yes
    toc_depth: 4
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    comment   = "#>",
    collapse  = TRUE,
    message   = FALSE,
    warning   = FALSE,
    fig.width = 7,
    fig.height = 5
)
```

# Overview

This vignette demonstrates how to analyze cell spatial graphs using
**igraph** functions directly. While CellGraphR provides import
(`importGraphH5`) and visualization (`plotGraph`) utilities, the igraph
package itself offers a rich set of graph algorithms for topology
analysis, community detection, and centrality computation.

We cover five practical analysis tasks:

1. **Subgraph extraction** --- spatial ROI and cell-type filtering
2. **Graph-level metrics** --- density, clustering, assortativity,
   connected components
3. **Community detection** --- Louvain algorithm and modularity
4. **Cell-type interface analysis** --- mixing matrices and boundary
   edges
5. **Node-level metrics** --- degree, local clustering, centrality

All examples use synthetic data so the vignette builds without external
files. Code blocks marked `eval=FALSE` show the equivalent operations
on real HDF5 graphs.

# Setup

```{r load-packages}
library(CellGraphR)
library(igraph)
library(ggplot2)
```

## Synthetic data

We generate a tissue-like spatial graph with 200 cells of four types,
spatially clustered to mimic real tissue architecture.

```{r synthetic-data}
set.seed(42L)
n_cells <- 200
cell_types <- c("neoplastic", "stromal", "inflammatory", "necrotic")

## Cluster centres for spatial realism
centres <- list(
    neoplastic   = c(250, 250),
    stromal      = c(750, 250),
    inflammatory = c(250, 750),
    necrotic     = c(750, 750)
)

nuclei <- lapply(seq_len(n_cells), function(i) {
    ct <- sample(cell_types, 1, prob = c(0.4, 0.3, 0.2, 0.1))
    ctr <- centres[[ct]]
    list(
        centroid = ctr + rnorm(2, sd = 150),
        type     = ct
    )
})

g <- buildSpatialGraph(nuclei, type = "delaunay")
```

```{r show-graph}
cat("Nodes:", vcount(g), "\n")
cat("Edges:", ecount(g), "\n")
cat("Vertex attributes:", paste(vertex_attr_names(g), collapse = ", "), "\n")
```

### Using real HDF5 data

```{r real-h5, eval=FALSE}
gz_file <- system.file(
    "extdata",
    "TCGA-EJ-5494-01Z-00-DX1.f466c17c-4671-45e0-aca9-0ceeec2c0280_graph.h5.gz",
    package = "CellGraphR"
)
h5_file <- tempfile(fileext = ".h5")
R.utils::gunzip(gz_file, destname = h5_file, remove = FALSE)
g_real <- importGraphH5(h5_file)
vertex_attr_names(g_real)
```

# Subgraph Extraction

## By spatial region of interest (ROI)

Extract the subgraph within a bounding box. This is useful for focusing
on a specific tissue region --- for example, the tumour core or the
invasion front.

```{r subgraph-roi}
## Get coordinate ranges
x_coords <- V(g)$x
y_coords <- V(g)$y
cat("X range:", range(x_coords), "\n")
cat("Y range:", range(y_coords), "\n")

## Define a bounding box (lower-left quadrant)
xmin <- min(x_coords); xmax <- median(x_coords)
ymin <- min(y_coords); ymax <- median(y_coords)

## Filter vertices within the box
in_roi <- which(x_coords >= xmin & x_coords <= xmax &
                y_coords >= ymin & y_coords <= ymax)

g_roi <- induced_subgraph(g, in_roi)
cat("Original:", vcount(g), "nodes,", ecount(g), "edges\n")
cat("ROI:     ", vcount(g_roi), "nodes,", ecount(g_roi), "edges\n")
```

## By cell type

Extract the subgraph containing only specific cell types. This isolates
intra-type connectivity patterns.

```{r subgraph-celltype}
## Keep only neoplastic and stromal cells
keep_types <- c("neoplastic", "stromal")
vids <- which(V(g)$type %in% keep_types)
g_sub <- induced_subgraph(g, vids)

cat("Retained:", vcount(g_sub), "nodes of types:",
    paste(keep_types, collapse = ", "), "\n")
cat("Edges:   ", ecount(g_sub), "\n")

## Cell-type composition in the subgraph
table(V(g_sub)$type)
```

# Graph-Level Metrics

igraph provides efficient, C-backed implementations of standard graph
metrics. All functions below scale well to graphs with millions of
nodes.

```{r graph-metrics}
metrics <- data.frame(
    metric = c(
        "Edge density",
        "Global transitivity",
        "Mean local clustering",
        "Assortativity (cell type)",
        "Assortativity (degree)",
        "Connected components",
        "Largest component fraction"
    ),
    value = c(
        round(edge_density(g), 5),
        round(transitivity(g, type = "global"), 4),
        round(transitivity(g, type = "localaverage"), 4),
        round(assortativity_nominal(g, factor(V(g)$type)), 4),
        round(assortativity_degree(g), 4),
        components(g)$no,
        round(max(components(g)$csize) / vcount(g), 4)
    )
)
knitr::kable(metrics, col.names = c("Metric", "Value"),
             caption = "Global graph-topology metrics.")
```

## Comparing metrics across cell-type subgraphs

```{r per-type-metrics}
type_metrics <- do.call(rbind, lapply(unique(V(g)$type), function(ct) {
    vids <- which(V(g)$type == ct)
    if (length(vids) < 3) return(NULL)
    gs <- induced_subgraph(g, vids)
    data.frame(
        cell_type    = ct,
        n_nodes      = vcount(gs),
        n_edges      = ecount(gs),
        density      = round(edge_density(gs), 5),
        transitivity = round(transitivity(gs, type = "global"), 4),
        n_components = components(gs)$no,
        stringsAsFactors = FALSE
    )
}))

knitr::kable(type_metrics,
             caption = "Graph metrics per cell-type subgraph.")
```

# Community Detection

Community detection identifies densely-connected subgroups in the graph.
The Louvain algorithm is a good default: it scales nearly linearly and
produces high-modularity partitions.

```{r community-detection}
comm <- cluster_louvain(g)

cat("Number of communities:", length(comm), "\n")
cat("Modularity:           ", round(modularity(comm), 4), "\n")
```

```{r community-composition}
## Community composition by cell type
comm_table <- table(
    community = membership(comm),
    cell_type = V(g)$type
)
knitr::kable(comm_table,
             caption = "Cell-type composition per community.")
```

```{r plot-communities, fig.cap="Graph coloured by Louvain community membership."}
## Add community membership as a vertex attribute for plotting
V(g)$community <- as.character(membership(comm))

plotGraph(g, color_by = "community",
          title = "Louvain communities")
```

### Alternative algorithms

```{r alt-community, eval=FALSE}
## Walktrap (random-walk based)
comm_wt <- cluster_walktrap(g)

## Fast greedy (hierarchical agglomeration)
## Note: can be slow on very large graphs (> 100k nodes)
comm_fg <- cluster_fast_greedy(g)

## Compare modularities
c(louvain    = modularity(comm),
  walktrap   = modularity(comm_wt),
  fast_greedy = modularity(comm_fg))
```

# Cell-Type Interface Analysis

## Mixing matrix

The mixing matrix counts edges between each pair of cell types.
Diagonal entries represent homotypic edges (within the same type);
off-diagonal entries represent heterotypic edges (between types).

```{r mixing-matrix}
## Get endpoint types for every edge
edge_ends <- ends(g, E(g), names = FALSE)
type_from <- V(g)$type[edge_ends[, 1]]
type_to   <- V(g)$type[edge_ends[, 2]]

## Build the mixing matrix
mix_counts <- table(type_from, type_to)

## Symmetrise (since the graph is undirected, (A,B) and (B,A) both appear)
mix_mat <- as.matrix(mix_counts)
mix_sym <- mix_mat + t(mix_mat)
diag(mix_sym) <- diag(mix_mat)  # diagonal was already correct

knitr::kable(mix_sym,
             caption = "Edge counts between cell-type pairs.")
```

```{r mixing-fractions}
## Normalise to edge fractions
mix_frac <- mix_sym / ecount(g)
knitr::kable(round(mix_frac, 4),
             caption = "Edge fractions between cell-type pairs.")
```

## Interface edges

Extract edges that connect two different cell types (heterotypic
edges). These represent the spatial interface between cell populations.

```{r interface-edges}
## Identify heterotypic edges
is_heterotypic <- type_from != type_to
cat("Total edges:      ", ecount(g), "\n")
cat("Heterotypic edges:", sum(is_heterotypic), "\n")
cat("Heterotypic ratio:", round(mean(is_heterotypic), 4), "\n")
```

### Interface subgraph between specific types

```{r interface-subgraph}
## Extract edges between neoplastic and inflammatory cells
is_interface <- (type_from == "neoplastic" & type_to == "inflammatory") |
                (type_from == "inflammatory" & type_to == "neoplastic")

cat("Neoplastic-inflammatory interface edges:", sum(is_interface), "\n")

## Build the interface subgraph
if (sum(is_interface) > 0) {
    g_interface <- subgraph.edges(g, which(is_interface))
    cat("Interface vertices:", vcount(g_interface), "\n")
}
```

# Node-Level Metrics

## Degree and local clustering

```{r node-metrics}
node_df <- data.frame(
    node_id   = V(g)$name,
    cell_type = V(g)$type,
    x         = V(g)$x,
    y         = V(g)$y,
    degree    = degree(g),
    clustering = transitivity(g, type = "local"),
    stringsAsFactors = FALSE
)

head(node_df)
```

```{r degree-by-type, fig.cap="Degree distribution by cell type."}
ggplot(node_df, aes(x = cell_type, y = degree, fill = cell_type)) +
    geom_boxplot(show.legend = FALSE) +
    labs(title = "Degree distribution by cell type",
         x = "Cell type", y = "Degree") +
    theme_bw()
```

## Per-cell-type summaries

```{r type-summaries}
type_summary <- aggregate(
    cbind(degree, clustering) ~ cell_type,
    data = node_df,
    FUN = function(x) round(mean(x, na.rm = TRUE), 4)
)
knitr::kable(type_summary,
             caption = "Mean degree and clustering by cell type.")
```

## Closeness centrality

`closeness()` is efficient for connected graphs. For disconnected
graphs, use `harmonic_centrality()` instead.

```{r closeness}
node_df$closeness <- closeness(g, normalized = TRUE)

## Top 5 most central nodes
top5 <- head(node_df[order(-node_df$closeness), ], 5)
knitr::kable(top5, row.names = FALSE,
             caption = "Top 5 nodes by closeness centrality.")
```

## Betweenness centrality

> **Performance note:** `betweenness()` is O(V * E), which is
> prohibitive for graphs with > 50k nodes. For large graphs, use
> `estimate_betweenness(cutoff = k)` to limit path-length
> consideration, or compute betweenness on a sampled subgraph.

```{r betweenness}
## Safe for our 200-node example
node_df$betweenness <- betweenness(g, normalized = TRUE)
```

```{r betweenness-large, eval=FALSE}
## For large graphs (> 50k nodes), use the cutoff variant:
btwn <- estimate_betweenness(g_real, cutoff = 3)
```

```{r betweenness-plot, fig.cap="Betweenness centrality by cell type."}
ggplot(node_df, aes(x = cell_type, y = betweenness, fill = cell_type)) +
    geom_boxplot(show.legend = FALSE) +
    labs(title = "Betweenness centrality by cell type",
         x = "Cell type", y = "Betweenness (normalised)") +
    theme_bw()
```

# Performance Tips for Large Graphs

When working with real histopathology graphs (~1M nodes, ~1.2M edges),
keep the following in mind:

| Function | Complexity | 1M-node feasibility |
|---|---|---|
| `vcount()`, `ecount()`, `edge_density()` | O(1) | Instant |
| `degree()` | O(V) | Instant |
| `transitivity(type="global")` | O(E * sqrt(E)) | Fast (seconds) |
| `components()` | O(V + E) | Fast |
| `cluster_louvain()` | ~O(V log V) | Fast (seconds) |
| `assortativity_nominal()` | O(E) | Fast |
| `closeness()` | O(V * (V + E)) | **Slow** --- minutes to hours |
| `betweenness()` | O(V * E) | **Infeasible** --- use `estimate_betweenness(cutoff=)` |
| `cluster_fast_greedy()` | O(V * E * log V) | **Slow** --- prefer Louvain |

**General strategy for large graphs:**

1. Use `induced_subgraph()` to extract a manageable ROI first
2. Prefer Louvain for community detection
3. Use `estimate_betweenness(cutoff = 3)` instead of exact betweenness
4. For closeness, compute on the largest connected component only, or
   use `harmonic_centrality()` which handles disconnected graphs

# Session Info

```{r session-info}
sessionInfo()
```
