---
title: "Visualizing Cell Spatial Graphs"
author: "Sehyun Oh"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteIndexEntry{Graph Visualization}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
output:
  BiocStyle::html_document:
    number_sections: yes
    toc: yes
    toc_depth: 4
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    comment   = "#>",
    collapse  = TRUE,
    message   = FALSE,
    warning   = FALSE,
    fig.width = 7,
    fig.height = 5,
    root.dir = "../"
)
```

# Overview

This vignette demonstrates how to use the two visualization functions
introduced in **CellGraphR**:

| Function | Purpose |
|----------|---------|
| `buildSpatialGraph()` | Convert a nuclei list into an `igraph` for one of five spatial graph types |
| `plotGraph()` | Render any `igraph` as a `ggplot2` figure with spatial or force-directed layout |

Together they let you inspect the raw cell-graph topology on top of the
actual tissue coordinates — before or alongside any feature extraction.

# Setup
```{r packages}
library(CellGraphR)
library(igraph)
library(ggplot2)
library(ImageFeatureTCGA)
library(SpatialExperiment)
```

# Simulated nuclei data {#sec-sim}

Both functions work with any `igraph` object, but `buildSpatialGraph()`
accepts the same `nuclei_list` format used by `extract_graph_features()`
— a list where each element has a `centroid` (numeric length-2 x/y
vector) and a `type` (character cell-type label).

Here we simulate 80 cells scattered across a 500 × 500 pixel field with
three cell types representative of a prostate tumour microenvironment.

```{r sim-data}
set.seed(123)
n_cells  <- 80
cell_types <- c("neoplastic", "stromal", "immune")

nuclei_list <- lapply(seq_len(n_cells), function(i) {
    list(
        centroid = runif(2, min = 0, max = 500),
        type     = sample(cell_types, size = 1,
                          prob = c(0.55, 0.30, 0.15))
    )
})

## Tabulate composition
table(sapply(nuclei_list, `[[`, "type"))
```

# Building spatial graphs {#sec-build}

`buildSpatialGraph()` constructs one of five graph representations from
the centroid coordinates. It returns an `igraph` with `x`, `y`, and
`type` vertex attributes — ready to pass straight to `plotGraph()`.

```{r build-all}
g_del <- buildSpatialGraph(nuclei_list, type = "delaunay")
g_knn <- buildSpatialGraph(nuclei_list, type = "knn", k = 5)
g_mst <- buildSpatialGraph(nuclei_list, type = "mst")
g_gab <- buildSpatialGraph(nuclei_list, type = "gabriel")
g_rng <- buildSpatialGraph(nuclei_list, type = "rng")

## Edge counts for each representation
data.frame(
    type  = c("Delaunay", "5-NN", "MST", "Gabriel", "RNG"),
    edges = c(ecount(g_del), ecount(g_knn), ecount(g_mst),
              ecount(g_gab), ecount(g_rng))
)
```

> **Graph type recap**
>
> | Type | Connectivity rule |
> |------|-------------------|
> | Delaunay | Adaptive distance threshold (20th-percentile cut-off); minimum degree ≥ 2 |
> | k-NN | Mutual k-nearest neighbours (edge requires both nodes to list each other) |
> | MST | Minimum spanning tree — the minimal connected subgraph (Prim's algorithm) |
> | Gabriel | Edge (i,j) if no third point lies inside the disc with diameter i–j |
> | RNG | Edge (i,j) if no third point lies in the lune between i and j |

# Plotting graphs {#sec-plot}

## Basic usage

`plotGraph()` auto-detects `x` and `y` vertex attributes for the
spatial layout and the `type` attribute for node colouring.

```{r plot-delaunay, fig.cap="Delaunay triangulation coloured by cell type. Nodes sit at real centroid coordinates."}
plotGraph(g_del)
```

## Customising the appearance

```{r plot-knn, fig.cap="Mutual 5-NN graph. Edges are sparser than the Delaunay graph because only reciprocal neighbours are connected."}
plotGraph(
    g_knn,
    color_by   = "type",
    title      = "Mutual 5-NN graph",
    node_size  = 2,
    edge_alpha = 0.4
)
```

## All five graph types

Plotting each type side by side makes the structural differences
immediately visible — from the dense Delaunay to the tree-like MST.

```{r plot-mst, fig.cap="Minimum spanning tree. Every cell is reachable via exactly n-1 edges."}
plotGraph(g_mst, title = "Minimum Spanning Tree",
          node_size = 2, edge_alpha = 0.6)
```

```{r plot-gabriel, fig.cap="Gabriel graph. A subset of the Delaunay triangulation defined by empty-disc criterion."}
plotGraph(g_gab, title = "Gabriel Graph", node_size = 2)
```

```{r plot-rng, fig.cap="Relative Neighborhood Graph. The sparsest proximity graph — a subgraph of the Gabriel graph."}
plotGraph(g_rng, title = "Relative Neighborhood Graph",
          node_size = 2, edge_alpha = 0.5)
```

# Working with real HDF5 graphs {#sec-h5}

When a slide is imported with `importGraphH5()`, the resulting `igraph`
carries the vertex attributes from the HDF5 `nodes` table — typically
spatial coordinates and a cell-type column. Pass the coordinate and
colour column names explicitly if auto-detection does not pick them up.

```{r h5-example, eval=FALSE}
graphFeatPath <- "TCGA-KK-A59Y-01Z-00-DX1.0DB9AF4D-4879-4295-A321-CBEE0FE6B2A0_graph.h5"
g <- importGraphH5(file.path(extDir, graphFeatPath))

## Inspect what vertex attributes are available
vertex_attr_names(g)

## Plot — replace "x", "y", "cell_type" with the actual column names
plotGraph(
    g,
    x_col    = "x",        # adjust to match your HDF5 schema
    y_col    = "y",
    color_by = "cell_type",
    node_size  = 0.5,      # small: thousands of cells
    edge_alpha = 0.1
)
```

> **Tip:** For slides with tens of thousands of cells the default node
> size of 1 can produce a dense cloud. Try `node_size = 0.3` and
> `edge_alpha = 0.05` for a cleaner look.

## Get coordination
```{bash eval=FALSE}
## Get the HoVer-Net output
## Under `inst/extdata/` directory
wget https://store.cancerdatasci.org/hovernet/h5ad/TCGA-KK-A59Y-01Z-00-DX1.0DB9AF4D-4879-4295-A321-CBEE0FE6B2A0.h5ad.gz
```

```{r}
# Read and convert to SingleCellExperiment
# extDir <- system.file("extdata", "CellGraphR")
extDir <- "~/Packages/CellGraphR/inst/extdata"
fpath <- "TCGA-KK-A59Y-01Z-00-DX1.0DB9AF4D-4879-4295-A321-CBEE0FE6B2A0.h5ad.gz"
dat <- ImageFeatureTCGA::HoverNet(file.path(extDir, fpath)) |> import()
```

```{r real-coords-overlay, fig.cap="Cell graph built from real HoverNet centroid coordinates overlaid on the tissue thumbnail."}
## obsm['spatial'] in the h5ad → reducedDim(dat, 'spatial') in the SCE
## Each row is a cell; row number == node number in the graph
coords     <- SpatialExperiment::spatialCoords(dat)  # n_cells × 2 (x, y)
cell_types <- colData(dat)$type   # adjust to match actual colData column name

## Whole-slide images can have > 1 M cells; subsample for practical visualisation
set.seed(42)
idx        <- sample(seq_len(nrow(coords)), min(10000L, nrow(coords)))
nuclei_sub <- lapply(idx, function(i)
    list(centroid = as.numeric(coords[i, ]), type = as.character(cell_types[i])))

## Build a Delaunay spatial graph from the subsampled cells
g_real <- buildSpatialGraph(nuclei_sub, type = "delaunay")

## Locate the matching tissue thumbnail
img_file <- file.path(extDir,
    "TCGA-KK-A59Y-01Z-00-DX1.0DB9AF4D-4879-4295-A321-CBEE0FE6B2A0.png")

## plotGraph() auto-scales coordinates to the image when graph space > image dims
plotGraph(
    g_real,
    image      = img_file,
    color_by   = "type",
    node_size  = 0.5,
    edge_alpha = 0.10,
    title      = "Real-coordinate cell graph on tissue"
)
```

# Overlaying graphs on tissue images {#sec-overlay}

`plotGraph()` can render the graph directly on top of the source
histopathology image using the `image` parameter. This section
demonstrates four common overlay use cases with the TCGA-KK-A59Y
example slide bundled in CellGraphR.

## Load the graph and locate the image

```{r overlay-setup, eval=FALSE}
h5_file <- system.file(
    "extdata",
    "TCGA-KK-A59Y-01Z-00-DX1.0DB9AF4D-4879-4295-A321-CBEE0FE6B2A0_graph.h5",
    package = "CellGraphR"
)
img_file <- system.file(
    "extdata",
    "TCGA-KK-A59Y-01Z-00-DX1.0DB9AF4D-4879-4295-A321-CBEE0FE6B2A0.png",
    package = "CellGraphR"
)

g <- importGraphH5(h5_file)
vertex_attr_names(g)
```

## Basic image overlay

Pass a file path to `image` and the tissue tile is rendered as the
background layer. Edge and node colours automatically adjust for
visibility on dark backgrounds.

```{r overlay-basic, eval=FALSE, fig.cap="Cell graph overlaid on the source tissue image."}
plotGraph(g, image = img_file, node_size = 0.5, edge_alpha = 0.15)
```

## Overlay with cell-type colouring

Specify `color_by` to colour nodes by a vertex attribute. This makes it
easy to see how different cell populations are distributed across the
tissue.

```{r overlay-celltype, eval=FALSE, fig.cap="Graph overlay with nodes coloured by cell type."}
plotGraph(
    g,
    image      = img_file,
    color_by   = "type",
    node_size  = 0.5,
    edge_alpha = 0.15
)
```

## Overlay with a pre-loaded raster

Instead of a file path, you can pass a raster array directly — useful
when you want to pre-process the image (e.g. crop, enhance contrast)
before plotting.

```{r overlay-raster, eval=FALSE, fig.cap="Graph overlay using a pre-loaded raster array."}
img_arr <- png::readPNG(img_file)
plotGraph(
    g,
    image      = img_arr,
    color_by   = "type",
    title      = "Pre-loaded raster overlay",
    node_size  = 0.5,
    edge_alpha = 0.15
)
```

## Overlay with coordinate scaling

When the image resolution differs from the graph coordinate space, use
`image_scale` to align them. For example, `image_scale = 0.5` halves
every vertex coordinate before overlaying.

```{r overlay-scale, eval=FALSE, fig.cap="Graph overlay with image_scale = 0.5 (coordinates halved)."}
plotGraph(
    g,
    image       = img_file,
    color_by    = "type",
    title       = "Scaled coordinates (0.5×)",
    node_size   = 0.5,
    edge_alpha  = 0.15,
    image_scale = 0.5
)
```

# Combining visualization and feature extraction {#sec-combined}

`buildSpatialGraph()` and `extract_graph_features()` accept the same
`nuclei_list`, so you can visualize the graph and extract features in
the same workflow without duplicating the centroid data.

```{r combined}
## 1. Build the graph for visualization
g_vis <- buildSpatialGraph(nuclei_list, type = "knn", k = 5)

## 2. Extract topology features (uses the same nuclei_list)
feats <- extract_graph_features(nuclei_list, image_dims = c(500, 500))

## Feature summary for the k-NN representation (k = 5)
unlist(feats$knn$k5)
```

```{r combined-plot, fig.cap="5-NN graph for the same cells whose features were extracted above."}
plotGraph(g_vis,
          title     = sprintf("5-NN graph  |  clustering coeff = %.3f",
                              feats$knn$k5$clustering_coefficient),
          node_size = 2)
```

# Session information {#sec-session}

```{r session}
sessionInfo()
```
